---
title: "Econ 265: \n Introduction to Econometrics"
subtitle: "Topic 6: Variants on MLR"
author: "Moshi Alam"
format:
  revealjs:
    theme: serif
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  echo: true
editor: visual
---

## Introduction {.smaller}

-   Data scaling
-   Standardized coefficients
-   Nonlinearities in X
-   Interaction terms
-   Average partial effects
-   Adjusted R-squared
-   Over fitting <!-- - Predicting $\hat{y}$ with results from regression of $f(y)$ on $x$ -->

## Data Scaling {.smaller}

```{r}
library(wooldridge); library(stargazer);
model1 <- lm(bwght ~ cigs + faminc, bwght) # bwght in grams
model2 <- lm(I(bwght/16) ~ cigs + faminc, bwght) # bwght in pounds
bwght$packs <- bwght$cigs/20 # 1 pack =  20 cigs
model3 <- lm(bwght ~ packs + faminc, bwght) # use packs instead of cigs
stargazer(model1, model2, model3, type = "text")
```

## Data Scaling {.smaller}

-   Coefficients in a regression model are sensitive to the units of measurement

<!-- :::{.incremental} -->

-   The population model: $bwght_i = \beta_0 + \beta_1 cigs_i + \beta_2 faminc + u_i$ follows MLR 1-5
-   $bwght_i/16 = \beta_0/16 + \beta_1/16 cigs_i + \beta_2/16 faminc + u_i$
    -   Coefficients scaled down by factor with which dependent variable is scaled
-   $bwhgt_i = \beta_0 + 20\beta_1 cigs_i/20 + \beta_2 faminc + u_i$
    -   which simplifies to $bwhgt_i = \beta_0 + 20\beta_1 packs_i + \beta_2 faminc + u_i$
    -   Coefficients scaled up by factor with which independent variable is scaled
-   What about SE? R\^2? <!-- ::: -->

## Standardized Coefficients {.smaller}

Instead of asking what happens to $y$ when $x$ changes by 1 unit, we ask what happens to $y$ when $x$ changes by 1 standard deviation

-   This is useful because:
    -   $x$ may be measured in different units in different datasets
    -   $x$ may be a composite index of several variables
    -   $x$ could be scaled up or down differently in different datasets. E.g. GPA
-   But s.d. effects are only useful when interpreted relative to the mean of $x$
    -   So the standardized $x_i$ is $x^* = (x_i - \bar{x})/s_x$
-   Thus from the population model: $y = \beta_0 + \beta_1 x_1 + \ldots + \beta_k x_k + u$
    -   The standardized model is: $y = \beta_0^* + \beta_1^* x_1^* + \ldots + \beta_k^* x_k^* + u^*$
        -   where $\beta_j^* =  \frac{s_{x_j}}{s_y}\beta_j$ for $j = 1, \ldots, k$
-   How would you derive this?
-   This is easy to implement in R using `scale()` function

## Standardized Coefficients {.smaller}

::::: columns
::: column
Scaled estimation:

```{r}
model4 <- lm(scale(bwght) ~ scale(cigs) + scale(faminc), bwght)
stargazer(model1, model4, type = "text", omit.stat = c("f", "ser"))
```
:::

::: column
Compute manually:

```{r}
model1$coef["cigs"]* sd(bwght$cigs)/sd(bwght$bwght)
model1$coef["faminc"]* sd(bwght$faminc)/sd(bwght$bwght)
```
:::
:::::

## More examples {.smaller}

price $=\beta_0+\beta_1$ nox $+\beta_2$ crime $+\beta_3$ rooms $+\beta_4$ dist $+\beta_5$ stratio $+u$

::::: columns
::: column
```{r}
library(wooldridge)
summary(lm(price ~ nox + crime + rooms + dist + stratio, hprice2))
```
:::

::: column
```{r}
summary(lm(scale(price) ~ scale(nox) + scale(crime) + scale(rooms) + scale(dist) + scale(stratio), hprice2))
```
:::
:::::

## Functional forms {.smaller}

Recall from lecture 1

| Specification | Change in x | Effect on y               |
|---------------|-------------|---------------------------|
| Level-level   | +1 unit     | \+$b_1$ units             |
| Level-log     | +1%         | \+$\frac{b_1}{100}$ units |
| Log-level     | +1 unit     | \+$(100 \times b_1)\%$    |
| Log-log       | +1%         | \+$b_1\%$                 |

## Nonlinearities in X {.smaller}

$$y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{1i}^2 + \beta_3 x_{2i} + u_i$$

$$\frac{\partial y}{\partial x_{1i}} = \beta_1 + 2\beta_2 x_{1i}$$

Discuss the nature of price change of houses with respect to number of rooms:

```{r}
lm(log(price) ~ log(nox) + log(dist) + rooms + I(rooms^2), hprice2)
```

::: incremental
-   Based on estimates how does price change as \# of rooms go up?
-   Increasing or decreasing rates?
-   Are there tipping points?
    -   $|\beta_1/2\beta_2|$ is the tipping point for $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{1i}^2 + \beta_3 x_{3i} + u_i$ when $x_{1i}$ changes
:::

## Interaction terms {.smaller}

::: incremental
-   $p =  \beta_0 + \beta_1 sqrft + \beta_2 bdrms + \beta_3 sqrft \times bdrms + \beta_4 bthrms +u$


-   We can formalize it with conditional expecations:

-   $E(p|sqrft, bdrms, bthrms) = \beta_0 + \beta_1 sqrft + \beta_2 bdrms + \beta_3 sqrft \times bdrms + \beta_4 bthrms$

-   $$\frac{\partial E(p|sqrft, bdrms, bthrms)}{\partial bdrms} = \beta_2 + \beta_3 sqrft$$

-   So at sqrft = $\bar{sqrft}$, the effect of bdrms on price is $\beta_2 + \beta_3 \bar{sqrft}$

-   Easy to obtain given $\hat{\beta}_2$ and $\hat{\beta}_3$ and $\bar{sqrft}$

-   Similarly in the previous slide $\frac{\partial E(y | x_1 = \bar{x_1}, x_2)}{\partial x_1} = \beta_1 + 2\beta_2 \bar{x_1}$
:::

```{r}

model <-  lm(price ~ sqrft + bdrms + I(sqrft * bdrms), data = hprice1)
 model$coefficients["bdrms"] + model$coefficients["I(sqrft * bdrms)"] * mean(hprice1$sqrft)
```



## Adjusted R-squared {.smaller}

- Recall Econ 160
-   Adjusted R-squared is a modified version of R-squared that adjusts for the number of predictors in the model.
-   Compare goodness of fit of models with different numbers of X's. 
$$\text{Adjusted-}R^2 = 1 - \left( \frac{(1 - R^2)(n - 1)}{n - k - 1} \right)$$
    $R^2$ is the R-squared value, $n$ is the \# observations and $k$ is the number of X's
    -   Penalizes the addition of unnecessary predictors
-   The F-test could be related to the adjusted R-squared. (next topic)
    -   Can be written into $UR$ and $R$ models for the F-test
    -   F-test is useful when models are *nested*



## Overfitting/overcontrolling {.smaller}

- Including too many controls can distort causal interpretation.  
- Example: Estimating the effect of **beer tax** on traffic fatalities ins tates:

$$
\text{fatalities}_s = \beta_0 + \beta_1 \text{tax}_s + \beta_2 \text{miles}_s + \beta_3 \text{percmale}_s + \beta_4 \text{perc16_21}_s + \dots
$$

- **Should we control for beer consumption (`beercons`)?**  
  - No, it mediates the effect of `tax` on fatalities.  
  - Controlling for it absorbs the policy impact.

---

## The Problem of Overcontrolling {.smaller}

- Overcontrolling can remove meaningful variation in key variables.  
- Example: Estimating the effect of **pesticides** on health costs:
  - Controlling for doctor visits **blocks part of the effect**.
- Another case: **School quality → earnings**:
  - If quality raises education, controlling for education **understates** its impact.
- **Takeaway:** Control variables should reflect **causal logic**, not just maximize $R^2$.


# Qualitative data

---

## Binary variables {.smaller}

::: incremental
-   Binary variables are {0,1} variables. E.g. `female`, `married`, `smoker`
- Example: Estimating the effect of **gender** on wages:
- $wage_i = \beta_0 + \delta_0 female_i + \beta_1 educ_i + u_i$
- $\delta_0=\mathrm{E}( wage_i \mid female_i =1, educ_i )-\mathrm{E}( wage_i \mid female_i =0, educ_i )$
- Difference in wages **between females and males**, holding education constant.
- If $\delta_0 < 0$, women earn less than men on average, given education.
  - Males are the base group ($\beta_0$ is their intercept)[draw]
- Including both `male` and `female` leads to perfect collinearity.
- Alternatively: $wage_i = \alpha_0 + \gamma_0 male_i + \beta_1 educ_i + u_i$
  - Here, females are the base group.
:::



## tests for wage discrimination {.smaller}

wage $=\beta_0+\delta_0$ female $+\beta_1$ educ $+\beta_2$ exper $+\beta_3$ tenure $+u$


```{r}
library(wooldridge)
summary( lm(wage ~ female + educ + exper + tenure, data = wage1))
```









## Multiple categories {.smaller}

Marital status and gender:
$$
log(wage_i) = \beta_0 + \beta_1 marriedmale_i + \beta_2 marriedfemale_i + \beta_3 unmarriedfemale_i \\
+ \beta_4 educ_i + \beta_5 exper_i + \beta_6 tenure_i + \beta_7 exper^2_i + \beta_8 tenure^2_i + u_i$$
 What is the base group?

```{r}
summary(lm(log(wage) ~ I(married*(1-female)) + I(married*female) + I((1-married)*female) + educ + exper + tenure + I(exper^2) + I(tenure^2), data = wage1) )
```
- What is the wage gap between married females and unmarried females?


## Interactions with dummy variables {.smaller}

What is the point of interacting: Marital status X gender?

$$
log(wage_i) = \beta_0 + \beta_1  female_i + \beta_2 married_i + \beta_3 married_i * female_i \\
+ \beta_4 educ_i + \beta_5 exper_i + \beta_6 tenure_i + \beta_7 exper^2_i + \beta_8 tenure^2_i + u_i$$

```{r}
summary(lm(log(wage) ~ female + married + I((married)*female) + educ + exper + tenure + I(exper^2) + I(tenure^2), data = wage1))
```


## Different slopes {.smaller}

- Do men and women have different returns to education?
-  $wage_i = \beta_0 + \beta_1 educ_i + \beta_2 female_i + \beta_3 educ_i * female_i + u_i$

-  $E(wage_i | educ_i, female_i = 1) - E(wage_i | educ_i, female_i = 0) = \beta_3 educ_i$
-  $\uparrow$ educ by 1 unit  $\uparrow$  wage by $\beta_1 + \beta_3$ for females and $\beta_1$ for males.
<!-- -  Returns to  $\uparrow$  1 year of educ  $\uparrow$  wages for females relative to males by $\beta_3$ -->
```{r}
summary(lm(log(wage) ~ educ + female + I(educ*female) + exper + expersq + tenure + tenursq, data = wage1))
```
- What is the gender gap at the average education level? 



## Different models for diff groups {.smaller}

$$
 GPA_i =\beta_0+\beta_1  SAT_i +\beta_2  hsrankperc_i +\beta_3  tothrs_i +u_i
$$

- For any of the slopes to depend on gender, we simply interact it with $female_i$, and include it 

- To test if the model is different between men and women, then we need a model where the intercept and all slopes can be different across the two groups

$$
\begin{aligned}
 GPA_i = & \beta_0+\beta_1  sat_i +\beta_2  hsperc_i   +\beta_3  tothrs_i \\
& +\delta_0  female_i  +\delta_1  female_i  *  sat_i +\delta_2  female_i  *  hsperc_i +\delta_3  female_i  *  tothrs_i  \\
& +u_i 
\end{aligned}
$$

## {.smaller}

:::: columns
::: {.column width="50%"}
```{r}
summary(lm(cumgpa ~ sat + hsperc + tothrs, data = gpa3))
```
::::
::: {.column width="50%"}
```{r}
summary(lm(cumgpa ~ sat + hsperc + tothrs + female + I(sat*female) + I(hsperc*female) + I(tothrs*female), data = gpa3)) 
```
:::
::::

## Heteroskedasticity to be covered by Zeyi in the lab {.smaller}